---
title: "EDA of Spanish electricity generation time series"
output:
  html_document:
    df_print: paged
---

#### Next steps

-   Comment on correlations. Perhaps plot together any pairs of variables that are particularly highly correlated or anti-correlated as examples
-   Comment on any long-term trends
-   Investigate seasonality
-   Improve plotting of interpolated regions

```{r libraries, message=FALSE, warning=FALSE, error=FALSE}
# Load libraries
library(RCurl)
library(tidyverse)
library(magrittr)
library(plotly)
library(imputeTS)
library(lubridate)
library(timetk)
library(ggcorrplot)
```

```{r echo=TRUE}
# Load data (relative file path) and view contents
df <- read.csv("../data/energy_dataset.csv")

# Formatting time variable to datetime object
# NB:  Time contains both UTC+1 and UTC+2 time-zones - date functions automatically pick this up and convert to UTC

df$time <- as_datetime(df$time)
head(df)
```

```{r}
dim(df) # Number of rows and columns
```

```{r}
summary(df) # Distribution and missingness for each variable
```

## Missing data

There are eight columns that contain no values (all NAs or zeroes). These are removed here.

```{r}
df <- df %>% 
  select(-c(generation.fossil.coal.derived.gas,
            generation.fossil.oil.shale,
            generation.fossil.peat,
            generation.geothermal,
            generation.marine,
            generation.hydro.pumped.storage.aggregated,
            generation.wind.offshore,
            forecast.wind.offshore.eday.ahead))
```

```{r}
colSums(is.na(df)) # Number of missing values for each remaining variable
```

16 out of 22 of the remaining variables each have a maximum of 36 missing values. Looking at where these occur, we see that most missing values occur across variables. Specifically, 90% of the missing values occur over only 18 rows. The remaining 10% are spread across 28 rows.

```{r}
# Rows where data is missing
df[!complete.cases(df),]
```

```{r}
# Number of NAs per row
missing_per_row <- rowSums(is.na(df[!complete.cases(df),]))
missing_per_row
sum(missing_per_row > 1)
sum(missing_per_row[missing_per_row > 1])/sum(missing_per_row)
```

We can visualise where in the series the missing values occur using the following function:

```{r}
ggplot_na_distribution(df$generation.biomass)
```

However, this only allows us to look at one variable at a time. There are other functions in the imputeTS package which can be useful for investigating missingness in a (univariate) time series.

We have decided to interpolate the missing values using linear interpolation. The only places where this might be an issue is where are two gaps of length 6 and 8. These will be visualised below.

```{r}
# Interpolate missing values using linear interpolation
df_clean <- df %>% mutate(across(generation.biomass:price.actual, .fns = na_interpolation, option = 'linear'))

# Verifying missing values have been filled
df_clean[!complete.cases(df_clean),]
```

Visualising interpolation

```{r}
# Creating a 'missing' flag to indicate imputation
df_clean$interp_biomass <- as.integer(is.na(df$generation.biomass))

ggplot(data = df_clean[90:150,]) + geom_line(aes(x=time, y = generation.biomass, colour = interp_biomass), show.legend = FALSE) + scale_color_gradient(low="black", high="red") + labs(title = "Example of linear interpolation (in red) applied to the dataset")
```


**What are we left with in df_clean?**

35,000 obs across 20 variables, hourly from start 2015 to end 2018:

-   Energy generation from 14 different sources

-   Forecast onshore wind and solar, one day ahead

-   Forecast and actual total electricity load

-   Forecast and actual electricity price

And in the other dataset, not yet loaded: hourly weather data over the same period for a number of Spanish cities



### Focusing on generation

```{r}
gen_data <- df_clean %>% 
  select(c(1, starts_with("generation")))

# Giving vars simpler names
names(gen_data) <- c("time",
                     "Biomass",
                    "Lignite",
                    "Gas",
                    "Hard Coal",
                    "Oil",
                    "Hydro Pumped",
                    "Hydro River",
                    "Hydro Reservoir",
                    "Nuclear",
                    "Other",
                    "Other Renewable",
                    "Solar",
                    "Waste",
                    "Wind Onshore")
```

```{r}
# Converting to long form
long_gen_data <- gen_data  %>% 
  pivot_longer(cols = -time, names_to = "generation_source")

head(long_gen_data)
```

Getting statistics over entire dataset for each source

```{r}
overall_stats <- long_gen_data %>%
  group_by(generation_source) %>%
  summarise(sum = sum(value, na.rm = TRUE),
            mean = mean(value, na.rm = TRUE),
            sd = sd(value, na.rm = TRUE),
            max = max(value, na.rm = TRUE)) %>%
  mutate(generation_source= fct_reorder(generation_source, desc(sum)))

sources_by_sum <- arrange(overall_stats, desc(sum))
sources_by_sum

ggplot(data = overall_stats) + geom_col(aes(x = generation_source, y = sum)) + theme(axis.text.x = element_text(angle=90)) + labs(title = "Total electricity generated by source")

sources_by_sum %>% pivot_longer(cols = c(sum, mean, sd, max), names_to = "stat", 
                                values_to = "value") %>%
  filter(stat != "sum") %>%
  ggplot() + geom_col(aes(x = generation_source, y = value, fill = stat), position = "dodge") + theme(axis.text.x = element_text(angle=90)) + labs(title = "Other overall hourly statistics of each source")
```

Using this information to re-factor the generation_source variable, so that plots will order the sources from most to least important.

```{r}
long_gen_data$generation_source <- factor(long_gen_data$generation_source, levels =  sources_by_sum$generation_source)
```

Smoothing the data on a daily scale

```{r}
gen_data_smoothed <- long_gen_data %>% 
  group_by(generation_source) %>% 
  summarise_by_time(time, .by = 'day', adjusted = mean(value)) 

head(gen_data_smoothed)
```

Looking at each series over time in isolation, with quarterly smoothing (This function is nice as it includes an interactive feature)

```{r}
gen_data_smoothed %>% 
  filter(generation_source %in% sources_by_sum$generation_source[1:6]) %>% 
  plot_time_series(time, adjusted, .facet_ncol = 3,
                   .smooth_period = "3 months",
                   .title= "Six biggest generation sources by day, with 3-month smoothing line")

gen_data_smoothed %>% 
  filter(generation_source %in% sources_by_sum$generation_source[7:14]) %>% 
  plot_time_series(time, adjusted, .facet_ncol = 3,
                   .smooth_period = "3 months",
                   .title= "Seven smallest generation sources by day, with 3-month smoothing")
```

##### Long-term trends

```{r}
long_gen_data %>% 
  group_by(generation_source) %>% 
  summarise_by_time(time, .by = '3 months', adjusted = mean(value)) %>% 
  ggplot() + 
  geom_area(aes(x = time, y = adjusted, fill = generation_source), stat = "identity")
```

```{r}
# Proportions
gen_data_props <- cbind(gen_data[1], prop.table(as.matrix(gen_data[-1]), margin = 1))

long_gen_data_props <- gen_data_props %>% 
  pivot_longer(cols = -time, names_to = "generation_source") 

# Ordering sources by importance
long_gen_data_props$generation_source <- factor(long_gen_data_props$generation_source, levels =  sources_by_sum$generation_source)

long_gen_data_props %>%
  group_by(generation_source) %>% 
  summarise_by_time(time, .by = '3 months', adjusted = mean(value)) %>% 
  ggplot() + 
  geom_area(aes(x = time, y = adjusted, fill = generation_source), stat = "identity") +
  theme_minimal()
  
```

```{r}
# Correlation plot. Sample size for computational efficiency.
df_corr <- gen_data[1:10000,2:ncol(gen_data)]

df_corr %>% cor() %>%
  ggcorrplot(.,
            hc.order = TRUE,
            type = "lower",
            lab = FALSE)
```

```{r}
# Create time variables
long_gen_data$year <- year(long_gen_data$time)
long_gen_data$month <- month(long_gen_data$time)
long_gen_data$day <- day(long_gen_data$time)
long_gen_data$hour <- hour(long_gen_data$time)

long_gen_data_props$year <- year(long_gen_data_props$time)
long_gen_data_props$month <- month(long_gen_data_props$time)
long_gen_data_props$day <- day(long_gen_data_props$time)
long_gen_data_props$hour <- hour(long_gen_data_props$time)
```

Seasonality within a day

```{r}
hourly_generation_plot_by_year <- long_gen_data %>%
  ggplot(aes(x = hour, y = value, fill = generation_source)) +
  geom_bar(stat = "identity")

hourly_generation_plot_by_year
```

```{r}
# Proportions
hourly_generation_plot_by_year_props <- long_gen_data_props %>%
  ggplot(aes(x = hour, y = value, fill = generation_source)) +
  geom_bar(stat = "identity")

hourly_generation_plot_by_year_props
```

Seasonality within a month

```{r}
daily_generation_plot_by_year <- long_gen_data %>%
  ggplot(aes(x = day, y = value, fill = generation_source)) +
  geom_bar(stat = "identity") +
  facet_wrap(~year) + theme(legend.position = "none") 

daily_generation_plot_by_year
```

```{r}
# Note: Similar problem to raw values
daily_generation_plot_by_year_props <- long_gen_data_props %>%
  ggplot(aes(x = day, y = value, fill = generation_source)) +
  geom_bar(stat = "identity") +
  facet_wrap(~year) + theme(legend.position = "none") 

daily_generation_plot_by_year_props
```

Seasonality within a year

```{r}
monthly_generation_plot <- long_gen_data %>%
  ggplot(aes(x = month, y = value, fill = generation_source)) +
  geom_bar(stat = "identity") +
  facet_wrap(~year)

monthly_generation_plot
```

```{r}
# Note: Shorter bars for shorter months as above
monthly_generation_plot_props <- long_gen_data_props %>%
  ggplot(aes(x = month, y = value, fill = generation_source)) +
  geom_bar(stat = "identity") +
  facet_wrap(~year)

monthly_generation_plot_props
```

```{r eval=FALSE, include=FALSE}
price_vs_forecast <- df_clean %>%
  select(., c(time, price.actual, price.day.ahead)) 

## Need to offset by a day before taking difference?
price_vs_forecast$difference <- price_vs_forecast$price.day.ahead - price_vs_forecast$price.actual

price_vs_forecast %>%
  ggplot() +
  geom_line(aes(x=time, y=difference)) +
  ylab("Difference between predicted and actual price")
```

```{r eval=FALSE, include=FALSE}
load_vs_forecast <- df_clean %>%
  select(., c(time, total.load.actual, total.load.forecast))

load_vs_forecast$difference <- load_vs_forecast$total.load.forecast - load_vs_forecast$total.load.actual

load_vs_forecast %>%
  ggplot() +
  geom_line(aes(x=time, y=difference)) +
  ylab("Difference between predicted and actual load")
```
